---
title: FileArray - A DelayedArray backend for filematrix
author:
- name: Malte Thodberg
  email: maltethodberg@gmail.com
output:
  BiocStyle::html_document:
    toc_float: yes
package: FileArray
vignette: >
  %\VignetteIndexEntry{User guide}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Introduction

The FileArray package implements a `r CRANpkg("DelayedArray")` for the
`r CRANpkg("filematrix")` R-packages for handling massive out-of-memory
matrices.

# Installation

Install from github with devtools:

```{r eval=FALSE}
devtools::install_github("MalteThodberg/FileArray")
```

Load the package using

# Creating a `FileArray`

There a multiple ways to create a FileArray from and R matrix?

```{r}
library(FileArray)
X <- matrix(rnorm(1000), ncol=10)
```

Using the the `realization` machinery by coercing the matrix to a
FileArray, which will generate a temporary file backing file:

```{r}
as(X, "FileArray")
```

If you want to specify the location of the backing file you can use use
`writeFileArray`

```{r}
my_filematrix <- tempfile()
writeFileArray(X, filepath = my_filematrix)
```

Logical and integer matrices are supported:

```{r}
as(X > 1, "FileArray")

Y <- round(X)
storage.mode(Y) <- "integer"
as(Y, "FileArray")
```

As are matrices with dimension names:

```{r}
rownames(X) <- sprintf("GENE_%i", seq_len(nrow(X)))
colnames(X) <- sprintf("SAMP_%i", seq_len(ncol(X)))
writeFileArray(X)
```

Alternatively you can directly use functions from `filematrix` and point
to the backing file afterwards.

```{r}
my_filematrix <- tempfile()
filematrix::fm.create.from.matrix(mat=X, filenamebase = my_filematrix)
FileArray(my_filematrix)
```

Large .txt files can be processed similarly using
`filematrix::fm.create.from.text.file` or large matrices can be created
by successively adding columns to a `filematrix` (see
<https://cran.r-project.org/web/packages/filematrix/vignettes/FM1_UsingFilematrices.html#appending-columns>)

# Manipulating `FileArray`s using delayed operations and DelayedMatrixStats

`FileArray`s are simply `DelayedArray` objects and can be manipulated as
such. The usual conventions for extracting data from matrix-like objects
work as expected:

```{r}
out <- as(X, "FileArray")
dim(out)
head(rownames(out))
head(out[,1])
```

We can also perform manipulations like subsetting and arithmetic. Note
that these operations do not affect the data in the filematrix backend;
rather, they are delayed until the values are explicitly required, hence
the creation of the `DelayedMatrix` object.

```{r}
out[1:5,1:5] 
out * 2
out %*% runif(ncol(out))
```

We can also calculate matrix summary state using the `r Biocpkg("DelayedMatrixStats")` package, which supports block-wise processing:

```{r}
library(DelayedMatrixStats)
colSums2(out)
rowSums2(out)
rowVars(out)
```

# Session information

```{r}
sessionInfo()
```
